Retrofitting Effect Handlers onto OCaml

### 2021-03-23 retro-concurrency.tex
Address more comments
Add points about callee-saved registers

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Installing an exception handler simply pushes the program counter (|pc|) of the
handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|) to the stack. After this, the current exception pointer is
updated to be the current stack pointer (|rsp|). This creates a linked-list of
exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
and jumps to the |pc| of the handler.

% Note |caml_call_ocaml| is |caml_start_program| in ocaml-multicore source

@@

In order to illustrate the implementation of handling effects, consider the
example presented in Figure~\ref{code:effimpl}. The layout of the program state
as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
performs effect |E| which is handled in the outer-most handler, and is
immediately resumed. The arrows between the fibers are parent pointers. At
position |p1|, |rsp| is at the top of the fiber |f|.

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| performs 100 million iterations of installing exception handlers
and returning with a value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro
benchmarks, we observed that padding tight loops with a few |nop| instructions,
which changes the loop alignment, makes the code up to 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\clearpage

@@

\if{0}
\subsection*{Dynamic Semantics}

@@

\begin{proof}
\end{proof}
\fi

@@

This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
the presence of multi-shot continuations, compiler optimisations need to be
made aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.

>>>>

The lack of callee-saved registers also makes exception handling fast. In the
absence of callee-saved registers, no registers need to be saved when entering
a |try| block. Similarly, no registers need to be restored when handling an
exception. Installing an exception handler simply pushes the program counter
(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
in |Caml_state|) to the stack. After this, the current exception pointer is
updated to be the current stack pointer (|rsp|). This creates a linked-list of
exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
and jumps to the |pc| of the handler.

@@

Similar to exception handlers, the lack of callee-saved registers in OCaml
benefits effect handlers. There is no register state to save when entering an
effect handler or performing an effect. Similarly, there is no register state
to restore when handling an effect or resuming a continuation. This fortuitous
design choice in stock OCaml has a significant impact towards fast switching
between fibers in Multicore OCaml.

In order to illustrate the runtime support for handling effects, consider the
example presented in Figure~\ref{code:effimpl}. The layout of the program state
as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
performs effect |E| which is handled in the outer-most handler, and is
immediately resumed. The arrows between the fibers are parent pointers. At
position |p1|, |rsp| is at the top of the fiber |f|.

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| performs 100 million iterations of installing exception handlers
and returning with a value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from Farvardin et al.~\cite{Farvardin20}. For
micro benchmarks, we observed that padding tight loops with a few |nop|
instructions, which changes the loop alignment, makes the code up to 15\%
faster. Hence, the difference in running times under 15\% may not be
statistically significant.

@@

\if{0}
\clearpage

@@

\subsection*{Dynamic Semantics}

@@

\begin{proof}
\end{proof}

@@

This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
the presence of multi-shot continuations, compiler optimisations need to be
made aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.
\fi
